---
title: "Evaluating the Big Mac Index as a Predictor for Inflation"
author: "Marco Chau, Julia Wong, Cindy Ly"
date: '2023-03-27'
output: 
  bookdown::pdf_document2:
    toc: true
    number_sections: true
abstract: "Insert abstract paragraph here."
bibliography: references.bib
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
library(tidyverse)
library(readr)
library(ggplot2)
library(knitr)
library(kableExtra)
```


### Load in Big Mac and inflation data from directory
```{r}
canada_data <- read.csv("data\\canada_big_mac_data.csv")
britain_data <- read.csv("data\\britain_big_mac_data.csv")
japan_data <- read.csv("data\\japan_big_mac_data.csv")
combined_data <- read.csv("data\\combined_big_mac_data.csv")
```

### Convert date columns from character class to date class
```{r}
canada_data$date <- as.Date(canada_data$date)
britain_data$date <- as.Date(britain_data$date)
japan_data$date <- as.Date(japan_data$date)
combined_data$date <- as.Date(combined_data$date)
```

\newpage
# Introduction

The Big Mac Index (BMI) was introduced by *The Economist* in 1986 as an informal means of measuring purchasing power parity (PPP) between currencies. Mathematically, the BMI is the price ratio of Big Macs between two countries. The index provided a more tangible method of determining the difference between the real purchasing power between two countries that exchange rate alone may not reflect. Although the BMI is an informal measurement, its application is rooted in PPP theory which states that the exchange rate between two currencies is proportional to the ratio of price levels in their respective countries. The use of the Big Mac as a tool for economic analysis was subsequently coined the term 'Burgernomics'. Discrepancies between the BMI and the exchange rate of two countries suggests that one currency is overvalued (or undervalued) compared to the other. Since its conception in 1986, the BMI has been written about and evaluated extensively by economists, academics, and journalists for its strengths and limitations as a measure for real exchange rates between two currencies.^[@Clements:2012] Since the evaluation of the BMI as a measure of PPP has taken the spotlight, there is a limited amount of literature that discusses the use of Burgernomics as a predictor for inflation rate within a country. The inflation rate measures the price changes in a predetermined basket of goods in a given economy and is meant to reflect changes in consumer purchasing power from year to year. This paper extends the use of the BMI as a validator of exchange rates between two currencies to a validator of reported inflation rates of a currency between two periods in time. 

|   In this paper we examine the differences between the inflation rate and the changes in the price of Big Macs in three countries: Canada, Japan, and the United Kingdom between 2000 - 2022. By examining differences between the inflation rate of Big Macs (which will be the Big Mac Index in the context of this paper) and the reported inflation rates of each country, we hope to determine whether the BMI can be an accurate predictor of the inflation rate, and whether reported inflation rates reflect the real changes in purchasing power from one year to the next within a country's borders. To evaluate our findings, we propose the following hypotheses:

**Null Hypothesis: The Big Mac Index is not an accurate predictor of the inflation rate.**\
**Alternative Hypothesis: The Big Mac Index is an accurate predictor of the inflation rate.**

|   In section \@ref(data) we discuss: the source of the Big Mac and inflation data, how the BMI is calculated, how local currencies are calculated, the strengths and limitations of our source and methodology for analysis, and assumptions we made prior to analysis. In this section, we also provide summaries of our datasets. In section \@ref(results) we show the results from evaluating Pearson's R and the linear regression of each country. Section \@ref(discussion) discusses the implications that the results in section \@ref(results) means for our hypotheses and the broader economic context.

# Data
## Sources
The price of Big Macs in local currencies is sourced from data provided by the *The Economist*^[@Economist:2023] and the exchange rate used to calculate the price of Big Macs in U.S. Dollars (USD) is provided by *Reuters*.^[@Reuters:2022] The CSV file that combines this data was downloaded from a GitHub repository.^[@Github:2022] We also added to the dataset by calculating the BMI, the price of Big Macs in USD, and the inflation rate in a given year. The Big Mac Index was calculated by finding the percentage change difference of prices in the local currency from one year to the next.  USD prices of Big Macs which was calculated by dividing the price of Big Macs in the local currency by the reported exchange rates during the given year. The inflation rate of Canada was sourced from the Bank of Canada's website.^[@Boc:2023] The inflation rate of the United Kingdom was sourced from the Office for National Statistics^[@National:2023] and the inflation rate of Japan was sourced from macrotrends.^[@macrotrends:2023]

## Overview of Data
### Summary of Canadian data
```{r}
summary(canada_data)
```
Top 10 rows of Canadian data
```{r}
kbl(head(select(canada_data, local_price, big_mac_index, inflation_rate, date)), booktabs = T, linesep = "")
```

```{r canada_price_fig}
# Create line graph to show price points for Canada
ggplot(canada_data) +
  aes(x = date, y = local_price) +
  geom_point(colour = "black") +
  geom_line(colour = "orange") +
  theme_minimal() +
  ggtitle("Price of Big Macs in Canada") +
  labs(x = "Date", 
       y = "Price in CAD") 

# Create line graph to show Big Mac Index for Canada
ggplot(canada_data) +
  aes(x = date, y = big_mac_index) +
  geom_point(colour = "black") +
  geom_line(colour = "orange") +
  theme_minimal() +
  ggtitle("Big Mac Index in Canada") +
  labs(x = "Date", 
       y = "Index") 

# Create line graph for inflation rate in Canada
ggplot(canada_data) +
  aes(x = date, y = inflation_rate) +
  geom_point(colour = "black") +
  geom_line(colour = "orange") +
  theme_minimal() +
  ggtitle("Inflation Rate in Canada") +
  labs(x = "Date", 
       y = "Inflation (%)") 
```

### Summary of British data
```{r}
summary(britain_data)
```

```{r britain_price_fig}
# Create line graph to show data points for Britain
ggplot(britain_data) +
  aes(x = date, y = local_price) +
  geom_point(colour = "black") +
  geom_line(colour = "blue") +
  theme_minimal() +
  ggtitle("Price of Big Macs in Britain") +
  labs(x = "Date", 
       y = "Price in British Pounds")

# Create line graph to show Big Mac Index for Britain
ggplot(britain_data) +
  aes(x = date, y = big_mac_index) +
  geom_point(colour = "black") +
  geom_line(colour = "blue") +
  theme_minimal() +
  ggtitle("Big Mac Index in Britain") +
  labs(x = "Date", 
       y = "Index") 

# Create line graph for inflation rate in Britain
ggplot(britain_data) +
  aes(x = date, y = inflation_rate) +
  geom_point(colour = "black") +
  geom_line(colour = "blue") +
  theme_minimal() +
  ggtitle("Inflation Rate in Britain") +
  labs(x = "Date", 
       y = "Inflation (%)") 
```

Top 10 rows of British data
```{r}
kbl(head(select(britain_data, local_price, big_mac_index, inflation_rate, date)), booktabs = T, linesep = "")
```

### Summary of Japanese data
```{r}
summary(japan_data)
```

Top 10 rows of Japanese data
```{r}
kbl(head(select(japan_data, local_price, big_mac_index, inflation_rate, date)), booktabs = T, linesep = "")
```

```{r japan_price_fig}
# Create line graph to show data points for Japan
ggplot(japan_data) +
  aes(x = date, y = local_price) +
  geom_point(colour = "black") +
  geom_line(colour = "red") +
  theme_minimal() +
  ggtitle("Price of Big Macs in Japan") +
  labs(x = "Date", 
       y = "Price in Yen")

# Create line graph to show Big Mac Index for Japan
ggplot(canada_data) +
  aes(x = date, y = big_mac_index) +
  geom_point(colour = "black") +
  geom_line(colour = "red") +
  theme_minimal() +
  ggtitle("Big Mac Index in Japan") +
  labs(x = "Date", 
       y = "Index") 

# Create line graph for inflation rate in Japan
ggplot(canada_data) +
  aes(x = date, y = inflation_rate) +
  geom_point(colour = "black") +
  geom_line(colour = "red") +
  theme_minimal() +
  ggtitle("Inflation Rate in Japan") +
  labs(x = "Date", 
       y = "Inflation (%)") 
```

### Summary of Combined data:
```{r}
summary(combined_data)
```

## Distributions
```{r}
# Distribution of independent variable: big_mac_price
ggplot(combined_data, aes(x=dollar_price)) +
    geom_histogram(binwidth=0.5, colour="black", fill="white") +
  labs(x = "Price of Big Macs in USD") +
  ggtitle("Distribution of Big Mac Prices")

# Density curve
ggplot(combined_data, aes(x=dollar_price)) + 
  geom_density() + 
  scale_y_continuous(labels = scales::comma) +
  labs (x = "Price of Big Macs in USD") +
  ggtitle("Density Curve of Big Mac Prices")
```


## Scatterplots
```{r}
# Scatter plot of Big Mac Index and Inflation in Canada
ggplot(canada_data) +
  aes(x = inflation_rate, y = big_mac_index) +
  geom_point(colour = "orange") +
  theme_minimal() +
  labs(x = "Inflation rate (%)", y = "Big Mac Index") +
  ggtitle("Scatter Plot of Canadian Big Mac Index and Inflation")
```

```{r}
# Scatter plot of Big Mac Index and Inflation in Britain
ggplot(britain_data) +
  aes(x = inflation_rate, y = big_mac_index) +
  geom_point(colour = "blue") +
  theme_minimal() +
  labs(x = "Inflation rate (%)", y = "Big Mac Index") +
  ggtitle("Scatter plot of British Big Mac Index and Inflation")
```

```{r}
# Scatter plot of Big Mac Index and Inflation in Japan
ggplot(japan_data) +
  aes(x = inflation_rate, y = big_mac_index) +
  geom_point(colour = "red") +
  theme_minimal() +
  labs(x = "Inflation rate (%)", y = "Big Mac Index") +
  ggtitle("Scatter plot of Japanese Big Mac Index and Inflation")
```

```{r}
# Scatter plot of Big Mac Index and Inflation in all three countries combined
ggplot(combined_data) +
  aes(x = inflation_rate, y = big_mac_index) +
  geom_point(colour = "black") +
  theme_minimal() +
  labs(x = "Inflation rate (%)", y = "Big Mac Index") +
  ggtitle("Scatter plot of Combined Big Mac Index and Inflation")
```

# Results
## Pearson's R
```{r}
# Calculate Pearson's r for Canadian Big Mac Index and Inflation Rate
cor(canada_data[-1,]$big_mac_index, canada_data[-1,]$inflation_rate)
```

```{r}
# Calculate Pearson's r for British Big Mac Index and Inflation Rate
cor(britain_data[-1,]$big_mac_index, britain_data[-1,]$inflation_rate)
```

```{r}
# Calculate Pearson's r for Japanese Big Mac Index and Inflation Rate
cor(japan_data[-1,]$big_mac_index, japan_data[-1,]$inflation_rate)
```

```{r}
# Calculate Pearson's r for Combined Big Mac Index and Inflation Rate
cor(combined_data[-c(1,24,47),]$big_mac_index, combined_data[-c(1,24,47),]$inflation_rate)
```

```{r}
#Conduct simple linear regression for Canada
# The variable before the ~ is always the dependent variable (outcome)
# The variables after the ~ are the independent variables
# Store regression results
canada_lin_reg <- lm(inflation_rate ~ big_mac_index, data = canada_data)

# Display the stored results
summary(canada_lin_reg)
```

```{r}
#Conduct simple linear regression for Britain
# The variable before the ~ is always the dependent variable (outcome)
# The variables after the ~ are the independent variables
# Store regression results
britain_lin_reg <- lm(inflation_rate ~ big_mac_index, data = britain_data)

# Display the stored results
summary(britain_lin_reg)
```

```{r}
#Conduct simple linear regression for Japan
# The variable before the ~ is always the dependent variable (outcome)
# The variables after the ~ are the independent variables
# Store regression results
japan_lin_reg <- lm(inflation_rate ~ big_mac_index, data = japan_data)

# Display the stored results
summary(japan_lin_reg)
```

```{r}
#Conduct simple linear regression for all countries combined
# The variable before the ~ is always the dependent variable (outcome)
# The variables after the ~ are the independent variables
# Store regression results
combined_lin_reg <- lm(inflation_rate ~ big_mac_index, data = combined_data)

# Display the stored results
summary(combined_lin_reg)
```

## Canada

## United Kingdom

## Japan

# Discussion


